Kaluza--Klein Cosmology from five-dimensional Lovelock--Cartan Theory

### 2016-11-14 KKosmology.tex
First changes after the Referee report!

In four dimensions, the Einstein--Hilbert action with cosmological constant is the most general theory which leads to second order field equations for the metric. In higher dimensions, however, particular combinations of higher order terms in the curvature can be added to the gravitational action, whose variation with respect to the metric yields also second order field equations. The most general theory in arbitrary dimensions, which preserves this feature of the four-dimensional Einstein--Hilbert, is named the Lanczos--Lovelock action~\cite{Lanczos:1938sf,*Lovelock:1971yv}. Such a theory, has no ghosts~\cite{Zumino:1985dp} and it has the same degrees of freedom as the Einstein--Hilbert Lagrangian in arbitrary dimensions~\cite{Henneaux:1990au}.

@@

The Riemannian five-dimensional Lanczos--Lovelock theory has been widely studied in the literature. For instance, exact wormhole solutions have been found in vacuum~\cite{Dotti:2006cp,*Dotti:2007az} and coupled with matter fields~\cite{Mehdizadeh:2015jra}. In higher order Lanczos--Lovelock models, this class of exact solutions have been reported in~\cite{Mehdizadeh:2015dta} and in the compactified theory with torsion~\cite{Canfora:2008ka}. Additionally, the compactification of higher Lanczos--Lovelock terms has been considered in Refs.~\cite{MuellerHoissen:1985mm,*MuellerHoissen:1989yv} and their cosmology in Refs.~\cite{MuellerHoissen:1985ij,Deruelle:1986iv,Deruelle:1989fj}. It is worth mentioning that, classically, wormholes must be supported by a kind of energy compatible with the cosmological hypothesis. Their existence in these extended models implies the presence of degrees of freedom that could provide an explanation for the exotic matter/energy abundance in the universe.

On the other hand, it is well-known that GR assume the torsion-free condition {\it a priori}. However, as E.~Cartan first considered, it is possible to take metricity and parallelism as truly independent concepts~\cite{Cartan1922,*Cartan1924,*Cartan1925}. Such kind of geometries are known as Riemann--Cartan and offer the natural framework for Poincar\'e gauge theories, where the torsion appears as the field strength of translations, sourced by the spin-current~\cite{Kibble:1961ba,Sciama:1962,Hehl:1976kj}.
%In such a framework, the vielbein is interpreted as the gauge connection of translations and the torsion as its field strength. The formulation of gravity as a gauge theory has major benefits in the problem of a unified picture of interactions, since it allows to interpret it as arising through a localization procedure, where the spacetime is a Riemann--Cartan manifold. 
%Due to this, several authors have relaxed the torsion-free condition since, as E. Cartan argued, there is no a fundamental reason to assume a torsion-free condition {\it a priori}. 
Moreover, the vacuum predictions of its simplest formulation---the Einstein--Cartan theory---holds the experimental tests of GR. The cosmological consequences of non-Riemannian geometries has been widely studied in the literature (for a review, see Ref.~\cite{Puetzfeld:2004yg}). %For instance, the pioneer works of Kopczy\'nski showed how the introduction of a spinning dust, acting as a source of the curvature and torsion, avoids the appearance of cosmological singularities~\cite{Kopczynski:1972,Kopczynski:1973}. In the context of the early evolution of the universe, the spin density can drive the inflationary phase within the Einstein--Cartan theory, along with the avoidance of singularities~\cite{Gasperini:1986mv}. It has been shown in Ref.~\cite{Shie:2008ms} that torsional degrees of freedom might produce the current phase of accelerated expansion of the universe. Additionally, the nonminimal coupling of the Euler characteristic form with a scalar field in Riemann--Cartan spacetimes, allows the torsional degrees of freedom to behave as an exotic induced-matter content, customary for the phase of accelerated expansion~\cite{Toloza:2013wi}.

@@

\begin{acknowledgments}
  The authors would like to thank A.~R.~Zerwekh, I.~Schmidt, S.~Kovalenko, N.~A.~Neill, C.~O.~Dib, O.~Miskovic, R.~Olea, R.~Gannouji, A.~Toloza, A.~Cisterna and J.~Zanelli for valuable comments on this work. \mbox{O.~C.-F.} has been partially supported by the PAI project No.~79140040 (CONICYT--Chile) and F.~R. by the Fondecyt project No.~1160423. The Centro Cient\'ifico Tecnol\'ogico de Valpara\'iso (CCTVal) is funded by the Chilean Government through the Centers of Excellence Basal Financing Program of CONICYT.
\end{acknowledgments}
%_-----------------------------------------------------------------------------------------

@@

\item[$\bm{u_\pm<0}$]
  \begin{align}
    \gamma(t)&=-\sqrt{-u_\pm}\coth\tau,\\
    \phi(t)&=\phi_0\sinh^2\tau\exp\left[\tfrac{\mp\sqrt{6\Delta}}{4\alpha_2u_\pm}\cosh^2\tau\right],\\
    a(t)&=\tfrac{a_0}{\sqrt{-u_\pm}}\cosh\tau\exp\left[\tfrac{\pm\sqrt{6\Delta}} {24\alpha_2u_\pm}\cosh^2\tau\right],\\
    f^2(t)&=-u_\pm\tfrac{k}{a_0^2}\cosh^{-2}\tau\exp\left[\tfrac{\mp\sqrt{6\Delta}}{12\alpha_2u_\pm}\cosh^2\tau\right]\notag\\
    &\quad -u_\pm\tanh^2\tau + \tfrac{3\alpha_1\pm\sqrt{6\Delta}}{6\alpha_2}.
  \end{align}
\end{description}
%%%%%%%%% BIBLIOGRAPHY %%%%%%%%%
%%\bibliographystyle{apsrev4-1}
\bibliography{References.bib}

>>>>

In four dimensions, the Einstein--Hilbert action with cosmological
constant is the most general theory which leads to second order field
equations for the metric. In higher dimensions, however, particular
combinations of higher order terms in the curvature can be added to
the gravitational action, whose variation with respect to the metric
yields also second order field equations. The most general theory in
arbitrary dimensions, which preserves this feature of the
four-dimensional Einstein--Hilbert, is named the Lanczos--Lovelock
action~\cite{Lanczos:1938sf,*Lovelock:1971yv}. Such a theory, has no
ghosts~\cite{Zumino:1985dp} and it has the same degrees of freedom as
the Einstein--Hilbert Lagrangian in arbitrary
dimensions~\cite{Henneaux:1990au}. It is worth mentioning that in the
Palatini approach, where the metric and the connection are considered
as independent fields, there are families of Lagrangians which yield
to second order field equations, and do not possess
ghosts~\cite{Olmo:2011uz}.

@@

The Riemannian five-dimensional Lanczos--Lovelock theory has been
widely studied in the literature. For instance, exact wormhole
solutions which violate no energy conditions have been found in
vacuum~\cite{Dotti:2006cp,*Dotti:2007az}, and coupled with matter
fields satisfying the weak energy
conditions~\cite{Mehdizadeh:2015jra}. In higher order
Lanczos--Lovelock models, this class of exact solutions have been
reported in~\cite{Mehdizadeh:2015dta} and in the compactified theory
with torsion~\cite{Canfora:2008ka}. Additionally, the compactification
of higher Lanczos--Lovelock terms has been considered in
Refs.~\cite{MuellerHoissen:1985mm,*MuellerHoissen:1989yv} and their
cosmology in
Refs.~\cite{MuellerHoissen:1985ij,Deruelle:1986iv,Deruelle:1989fj}. It
is worth mentioning that classically, in GR, wormholes must be supported by
a kind of energy compatible with the cosmological hypothesis. Their
existence in these extended models implies the presence of degrees of
freedom that could provide an explanation for the exotic matter/energy
abundance in the universe.

On the other hand, it is well-known that GR assume the torsion-free
condition {\it a priori}. However, as E.~Cartan first considered, it
is possible to take metricity and parallelism as truly independent
concepts~\cite{Cartan1922,*Cartan1924,*Cartan1925}. Such kind of
geometries are known as Riemann--Cartan and offer the natural
framework for Poincar\'e gauge theories, where the torsion appears as
the field strength of translations, sourced by the
spin-current~\cite{Kibble:1961ba,Sciama:1962,Hehl:1976kj}.
%In such a framework, the vielbein is interpreted as the gauge connection of translations and the torsion as its field strength. The formulation of gravity as a gauge theory has major benefits in the problem of a unified picture of interactions, since it allows to interpret it as arising through a localization procedure, where the spacetime is a Riemann--Cartan manifold. 
%Due to this, several authors have relaxed the torsion-free condition since, as E. Cartan argued, there is no a fundamental reason to assume a torsion-free condition {\it a priori}. 
Moreover, the vacuum predictions of its simplest formulation---the
Einstein--Cartan theory---holds the experimental tests of GR. In the
same spirit of the Gauss--Bonnet extension of GR, the quadratic
corrections in curvature and torsion to Einstein--Cartan have been
considered, see for example Refs.~\cite{Baekler:2011jt,Fabbri:2012qr}.
The cosmological consequences of non-Riemannian geometries has been
widely studied in the literature (for a review, see
Ref.~\cite{Puetzfeld:2004yg}).
%For instance, the pioneer works of Kopczy\'nski showed how the introduction of a spinning dust, acting as a source of the curvature and torsion, avoids the appearance of cosmological singularities~\cite{Kopczynski:1972,Kopczynski:1973}. In the context of the early evolution of the universe, the spin density can drive the inflationary phase within the Einstein--Cartan theory, along with the avoidance of singularities~\cite{Gasperini:1986mv}. It has been shown in Ref.~\cite{Shie:2008ms} that torsional degrees of freedom might produce the current phase of accelerated expansion of the universe. Additionally, the nonminimal coupling of the Euler characteristic form with a scalar field in Riemann--Cartan spacetimes, allows the torsional degrees of freedom to behave as an exotic induced-matter content, customary for the phase of accelerated expansion~\cite{Toloza:2013wi}.

@@

\begin{acknowledgments}
  The authors would like to thank A.~R.~Zerwekh, I.~Schmidt,
  S.~Kovalenko, N.~A.~Neill, C.~O.~Dib, O.~Miskovic, R.~Olea,
  R.~Gannouji, A.~Toloza, A.~Cisterna and J.~Zanelli for valuable
  comments on this work. \mbox{O.~C.-F.} has been partially supported
  by the PAI project No.~79140040 (CONICYT--Chile) and F.~R. by the
  Fondecyt project No.~1160423. The Centro Cient\'ifico Tecnol\'ogico
  de Valpara\'iso (CCTVal) is funded by the Chilean Government through
  the Centers of Excellence Basal Financing Program FS0821 of CONICYT.
\end{acknowledgments}
%_-----------------------------------------------------------------------------------------

@@

\item[$\bm{u_\pm<0}$]
  \begin{align}
    \gamma(t)&=-\sqrt{-u_\pm}\coth\tau,\\
    \phi(t)&=\phi_0\sinh^2\tau\exp\left[\tfrac{\mp\sqrt{6\Delta}}{4\alpha_2u_\pm}\cosh^2\tau\right],\\
    a(t)&=\tfrac{a_0}{\sqrt{-u_\pm}}\cosh\tau\exp\left[\tfrac{\pm\sqrt{6\Delta}} {24\alpha_2u_\pm}\cosh^2\tau\right],\\
    f^2(t)&=-u_\pm\tfrac{k}{a_0^2}\cosh^{-2}\tau\exp\left[\tfrac{\mp\sqrt{6\Delta}}{12\alpha_2u_\pm}\cosh^2\tau\right]\notag\\
    &\quad -u_\pm\tanh^2\tau + \tfrac{3\alpha_1\pm\sqrt{6\Delta}}{6\alpha_2}.
  \end{align}
\end{description}
%%%%%%%%% BIBLIOGRAPHY %%%%%%%%%
%% \bibliographystyle{apsrev4-1}
\bibliography{References}
